---
title: "One page briefing"
output: html_notebook
---

# Does Scotland deserve its reputation as the 'sick man of Europe'? 

## It depends where you mean by 'Europe' *(But the answer's probably 'yes')*

Text goes here


```{r}
pacman::p_load(
  tidyverse
)

source("scripts/smoother_function.R")
source("scripts/scotland_in_context__helper_functions.R")

dta <- read_csv("data/tidy/new_counts.csv")

# country_code group definitions
source("scripts/scotland_in_context__country_group_definitions.R")

smooth_fn <- function(DTA, SMOOTH_PAR = 1.3){
  out <- DTA %>%   
    mutate(
      cmr = deaths / exposure,
      lg_cmr = log(cmr, base=10)     
    ) %>% smooth_var(
      dta =.,
      group_vars= c("country_code", "sex"),
      smooth_var = "lg_cmr",
      smooth_par = SMOOTH_PAR
    ) 
  return(out)
}


```


```{r}
dta_ne <- dta %>% 
  group_by(country_code, year, age) %>% 
  summarise(
    deaths = sum(deaths),
    population = sum(population),
    exposure = sum(exposure)
            ) %>% 
  ungroup() %>% 
  filter(
    country_code %in% europe_northern,
    year >= 1950 & year <= 2010
  ) %>% 
  group_by(
    year, age
  ) %>% 
  summarise(
    deaths = sum(deaths, na.rm = T), 
    exposure = sum(exposure, na.rm=T)
  ) %>% 
  ungroup() %>% 
  mutate(
    death_rate = deaths / exposure,
    lg_death_rate = log(death_rate, base = 10)
  )

dta_se  <- dta %>% 
  group_by(country_code, year, age) %>% 
  summarise(
    deaths = sum(deaths),
    population = sum(population),
    exposure = sum(exposure)
            ) %>% 
  ungroup() %>% 
  filter(
    country_code %in% europe_southern,
    year >= 1950 & year <= 2010
  ) %>% 
  group_by(
    year, age
  ) %>% 
  summarise(
    deaths = sum(deaths, na.rm = T), 
    exposure = sum(exposure, na.rm = T)
  ) %>% 
  ungroup() %>% 
  mutate(
    death_rate = deaths / exposure,
    lg_death_rate = log(death_rate, base = 10)
  )


dta_ee  <- dta %>% 
  group_by(country_code, year, age) %>% 
  summarise(
    deaths = sum(deaths),
    population = sum(population),
    exposure = sum(exposure)
            ) %>% 
  ungroup() %>% 
  filter(
    country_code %in% europe_eastern,
    year >= 1950 & year <= 2010
  ) %>% 
  group_by(
    year, age
  ) %>% 
  summarise(
    deaths = sum(deaths, na.rm = T), 
    exposure = sum(exposure, na.rm = T)
  ) %>% 
  ungroup() %>% 
  mutate(
    death_rate = deaths / exposure,
    lg_death_rate = log(death_rate, base = 10)
  )

dta_rwe  <- dta %>% 
  filter(
    country_code %in% europe_western[europe_western != "GBR_SCO"],
    year >= 1950 & year <= 2010
  ) %>% 
  group_by(country_code, year, age) %>% 
  summarise(
    deaths = sum(deaths),
    population = sum(population),
    exposure = sum(exposure)
            ) %>% 
  ungroup() %>% 
  group_by(
    year, age
  ) %>% 
  summarise(
    deaths = sum(deaths, na.rm = T), 
    exposure = sum(exposure, na.rm = T)
  ) %>% 
  ungroup() %>% 
  mutate(
    death_rate = deaths / exposure,
    lg_death_rate = log(death_rate, base = 10)
  )

dta_scot <- dta %>% 
  filter(country_code == "GBR_SCO") %>% 
    group_by(country_code, year, age) %>% 
  summarise(
    deaths = sum(deaths),
    population = sum(population),
    exposure = sum(exposure)
            ) %>% 
  ungroup() %>% 
  mutate(death_rate = deaths / exposure,
         lg_death_rate = log(death_rate, base = 10)
  )


# Difference between Scot and rwe
tmp1 <- dta_scot %>% 
  select(year, age, `Scotland` = lg_death_rate)  

tmp2 <- dta_rwe  %>% 
  select(year, age, `Rest of Western Europe` = lg_death_rate) 

dif_scot_rwe <- tmp1 %>% 
  left_join(tmp2) %>% 
  mutate(
    `Scot cf rest Western Europe` = `Scotland` - `Rest of Western Europe`
  ) %>% 
  select(
    year, age, `Scot cf rest Western Europe`
  )

rm(tmp1, tmp2)



# Difference between Scot and ne
tmp1 <- dta_scot %>% 
  select(year, age, `Scotland` = lg_death_rate)  

tmp2 <- dta_rwe  %>% 
  select(year, age, `Northern Europe` = lg_death_rate) 

dif_scot_ne <- tmp1 %>% 
  left_join(tmp2) %>% 
  mutate(
    `Scot cf Northern Europe` = `Scotland` - `Northern Europe`
  ) %>% 
  select(
    year, age, `Scot cf Northern Europe`
  )

rm(tmp1, tmp2)


# Difference between Scot and ne
tmp1 <- dta_scot %>% 
  select(year, age, `Scotland` = lg_death_rate)  

tmp2 <- dta_ee  %>% 
  select(year, age, `Eastern Europe` = lg_death_rate) 

dif_scot_ee <- tmp1 %>% 
  left_join(tmp2) %>% 
  mutate(
    `Scot cf Eastern Europe` = `Scotland` - `Eastern Europe`
  ) %>% 
  select(
    year, age, `Scot cf Eastern Europe`
  )

rm(tmp1, tmp2)

# Difference between Scot and se
tmp1 <- dta_scot %>% 
  select(year, age, `Scotland` = lg_death_rate)  

tmp2 <- dta_se  %>% 
  select(year, age, `Southern Europe` = lg_death_rate) 

dif_scot_se <- tmp1 %>% 
  left_join(tmp2) %>% 
  mutate(
    `Scot cf Southern Europe` = `Scotland` - `Southern Europe`
  ) %>% 
  select(
    year, age, `Scot cf Southern Europe`
  )

rm(tmp1, tmp2)


# Join the above together

comparisons <- dif_scot_rwe %>% 
  left_join(dif_scot_ne) %>% 
  left_join(dif_scot_se) %>% 
  left_join(dif_scot_ee) %>% 
  gather(key = "comparison", value = "dif_logs", -year, -age)


# 
# png(filename="figures/rostock/figure_02_clps.png", 
#     width=60, height=30, res=300, units="cm"
# )
# 
# comparisons %>% 
#   mutate(
#     dif_logs = ifelse(dif_logs < -0.40, -0.40, dif_logs),
#     dif_logs = ifelse(dif_logs > 0.40, 0.40, dif_logs)
#   ) %>%   
#   smooth_var(., 
#              group_vars = c("sex", "comparison"), 
#              smooth_var = "dif_logs", 
#              smooth_par = 0.7
#   ) %>% 
#   levelplot(
#     dif_logs ~ year * age | comparison + sex,
#     data=., 
#     region=T,
#     ylab="Age in years",
#     xlab="Year",
#     at = seq(from= -0.40, to = 0.40, by=0.01),
#     col.regions = colorRampPalette(rev(brewer.pal(6, "RdBu")))(200),
#     scales=list(alternating=3),
#     main=NULL,
#     aspect= "iso",
#     ylim=c(0, 90),
#     xlim=c(1950, 2010),
#     par.settings=list(strip.background=list(col="lightgrey"))
#   )
# 
# dev.off()

# Produce 'plain' version of above for adding annotations to  
# png(filename="figures/scotland_in_context/final_figures/figure_02a_clp_scot_uk_we_plain.png", 
#     width=30, height=30, res=300, units="cm"
# )
# 
# comparisons %>% 
#   mutate(
#     dif_logs = ifelse(dif_logs < -0.30, -0.30, dif_logs),
#     dif_logs = ifelse(dif_logs > 0.30, 0.30, dif_logs)
#   ) %>%   
#   smooth_var(., 
#              group_vars = c("sex", "comparison"), 
#              smooth_var = "dif_logs", 
#              smooth_par = 0.7
#   ) %>% 
#   levelplot(
#     dif_logs ~ year * age | comparison + sex,
#     data=., 
#     region=T,
#     ylab="Age in years",
#     xlab="Year",
#     at = seq(from= -0.30, to = 0.30, by=0.01),
#     col.regions = colorRampPalette(rev(brewer.pal(6, "RdBu")))(200),
#     scales=list(alternating=3),
#     main=NULL,
#     aspect= "iso",
#     ylim=c(0, 90),
#     xlim=c(1950, 2010),
#     par.settings=list(strip.background=list(col="lightgrey")) 
#   )

dev.off()


```

```{r produce_clp}
comparisons %>% 
  filter(year >= 1950) %>% 
  filter(year <= 2010) %>% 
  mutate(
         dif_logs = ifelse(dif_logs < -0.30, -0.30, dif_logs),
         dif_logs = ifelse(dif_logs > 0.30, 0.30, dif_logs)
  ) %>% 
  filter(age <= 90) %>% 
  mutate(comparison = stringr::str_replace(comparison, "Scot cf ", "")) %>%
  mutate(comparison = stringr::str_replace(comparison, " Europe", "")) %>%  
  ggplot(aes(x = year, y = age, fill = dif_logs)) + 
  geom_tile() + 
  facet_wrap(~comparison) +   
  scale_fill_gradientn(
    "",
    colours = rev(scales::brewer_pal(palette = "RdBu")(5)),
    limits = c(-0.301, 0.301)
  ) + 
  coord_fixed() + 
  guides(fill = F) + 
  geom_abline(slope = 1, intercept = seq(-2010, -1800, by = 10), linetype = "dashed", alpha = 0.5) + 
  theme_minimal() + 
  geom_vline(xintercept = seq(1950, 2010, by = 10), linetype = "dashed", alpha = 0.5) + 
  geom_hline(yintercept = seq(0, 90, by = 10), linetype = "dashed", alpha = 0.5) + 
  labs(x = "Year", y = "Age in single years", title = "Relative mortality risk for Scottish populations", subtitle = "Red means higher mortality in Scotland; darker shades are bigger differences", caption = "Source: Human Mortality Database") + 
  scale_y_continuous(sec.axis = sec_axis(~ 2010 - ., name = "Birth Cohort")) +
  theme(axis.text.y.right = element_text(angle = 45, hjust = -0.5))


```

Now difference in life expectancy 

```{r dif_e0}
e0_ee <- dta_ee %>% group_by(year) %>% mutate(da = age * deaths) %>% summarise(Eastern = sum(da) / sum(deaths))

e0_rwe <- dta_rwe %>% group_by(year) %>% mutate(da = age * deaths) %>% summarise(Western = sum(da) / sum(deaths))

e0_se <- dta_se %>% group_by(year) %>% mutate(da = age * deaths) %>% summarise(Southern = sum(da) / sum(deaths))

e0_ne <- dta_ne %>% group_by(year) %>% mutate(da = age * deaths) %>% summarise(Northern = sum(da) / sum(deaths))

e0_scot <- dta_scot %>% group_by(year) %>% mutate(da = age * deaths) %>% summarise(Scotland = sum(da) / sum(deaths))

e0_all <- e0_ee %>%
  left_join(e0_rwe) %>% 
  left_join(e0_se) %>% 
  left_join(e0_ne) %>% 
  left_join(e0_scot) %>% 
  gather(key = "Population", value = "e0", -year)



```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
